---
output: github_document
always_allow_html: true
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  fig.retina = 2,
  fig.align = 'center'
)
```

# portawaterperu

<!-- badges: start -->

[![License: CC BY
4.0](https://img.shields.io/badge/License-CC_BY_4.0-lightgrey.svg)](https://creativecommons.org/licenses/by/4.0/)

<!-- badges: end -->

Depending on how and what you would like to use the data for, the
description might change. It is based on the student project of Lauren
Judah..

## Installation

You can install the development version of portawaterperu from
[GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("openwashdata/portawaterperu")
```

```{r}
## Run the following code in console if you don't have the packages
## install.packages(c("dplyr", "knitr", "readr", "stringr", "gt", "kableExtra"))
library(dplyr)
library(knitr)
library(readr)
library(stringr)
library(gt)
library(kableExtra)
```

Alternatively, you can download the individual datasets as a CSV or XLSX
file from the table below.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

extdata_path <- "https://github.com/openwashdata/portawaterperu/raw/main/inst/extdata/"

read_csv("data-raw/dictionary.csv") |> 
  distinct(file_name) |> 
  dplyr::mutate(file_name = str_remove(file_name, ".rda")) |> 
  dplyr::rename(dataset = file_name) |> 
  mutate(
    CSV = paste0("[Download CSV](", extdata_path, dataset, ".csv)"),
    XLSX = paste0("[Download XLSX](", extdata_path, dataset, ".xlsx)")
  ) |> 
  knitr::kable()

```

## Data

The package provides access to ...

```{r}
library(portawaterperu)
```

### portawaterperu

The dataset `portawaterperu` contains data about ... It has
`r nrow(portawaterperu)` observations and `r ncol(portawaterperu)`
variables

```{r}
portawaterperu |> 
  head(3) |> 
  gt::gt() |>
  gt::as_raw_html()
```

For an overview of the variable names, see the following table.

```{r echo=FALSE, message=FALSE, warning=FALSE}
readr::read_csv("data-raw/dictionary.csv") |>
  dplyr::filter(file_name == "portawaterperu.rda") |>
  dplyr::select(variable_name:description) |> 
  knitr::kable() |> 
  kableExtra::kable_styling("striped") |> 
  kableExtra::scroll_box(height = "200px")
```

## Example

```{r}
library(portawaterperu)
library(ggplot2)
# Provide some example code here
portawaterperu |> 
  #dplyr::filter(stringr::str_starts(divisiones, "AMAZONAS")) |>
  #dplyr::group_by(divisiones) |> 
  #dplyr::summarise(mean = mean(pob_servida)) |> 
  ggplot(aes(y = pob_servida, color = tipo_gravedad))+
  geom_boxplot(outliers = F)+
  labs(title = "Population served given different gravity types",
       y= "Population") +
  theme_classic()
  
```

Refer to Laurens student project for nice examples. Here report
index.file is in the data-raw/report-lauren folder...

You will have to adjust the code slightly...

## License

Data are available as
[CC-BY](https://github.com/openwashdata/portawaterperu/blob/main/LICENSE.md).

## Citation

Please cite this package using:

```{r}
citation("portawaterperu")
```
